---
title: "HW4 Propensity Scores and IPTW: Answers"
authors: "Wilam and Shen"
output: pdf_document
---

**We are both submitting this assignment on NYU Classes.**

```{r, echo=FALSE, message=FALSE}
### Packages
require(arm)
require("Hmisc")
require(MatchIt)
require(optmatch)

```

### Question 1: Load the data and choose confounders (Step 1)
```{r}
### Answer 1
load("hw4.Rdata")
hw4 <- subset(hw4, bw<3000)
#remove white and college to avoid multicolinearity, don't add momed and state indicators
df <- data.frame(ppvtr.36=hw4$ppvtr.36, treat=hw4$treat, subset(hw4, select=c(bw, bwg, hispanic, black, b.marr, lths, hs, ltcoll, work.dur, prenatal, booze, cig, sex, first, preterm, momage, dayskidh, income)))
covariates <- colnames(df)[3:length(colnames(df))]
covariates


```

```{r}
### Answer 2
fit <- glm(treat ~ bw + bwg + hispanic + black + b.marr + lths + hs +
ltcoll + work.dur + prenatal + booze + cig + sex + first + preterm + momage + dayskidh + income,
family=binomial(link="logit"), data=df)
#pscores <- predict(fit, type='response') #same as line below
pscores <- fit$fitted.values
df$pscores <- pscores #unnecessary

```

```{r}
### Answer 3(a): We are interested in ATT 
# because the research question asks about the children that did receive treatment. 
```

```{r}
### Answer 3(b)
matches <- matching(z=df$treat, score=pscores, replace=TRUE)
weight <- c(rep(1,sum(df$treat==1)),matches$cnts)
#weight <- ifelse(df$treat == 0, matches$cnts, 1)

#saving weight as a unique variable for later
weight1 <- weight
```

```{r}
### Answer 4(a)

require(ggplot2)
plot <- ggplot() + geom_histogram(aes(subset(df, treat==1)$pscores), color="red", alpha=0) + geom_histogram(aes(subset(df, treat==0)$pscores), color="green", alpha=0) + ylim(0, 50)
plot

plot <- ggplot(data = df, aes(x = dayskidh)) + geom_density(data = subset(df, treat == 1), fill = "red", alpha = 0.7) + geom_density(data = subset(df, treat == 0), fill = "green", alpha = 0.7) + ylim(0,0.05)
plot

plot <- ggplot(data = df, aes(x = bw)) + geom_density(data = subset(df, treat == 1), fill = "red", alpha = 0.7) + xlim(1250, 3000) + geom_density(data = subset(df, treat == 0), fill = "green", alpha = 0.7) 
plot
```


```{r}
### Answer 4(b)
# The overlap on pscores is not perfect. Even though 
# there exist "empirical counterfactuals" for all
# treated when pscore = 1 we will need to match many 
# treated to the same control unit. In control group,
# kids stay in average less time in the hospital, but 
# there are enough units to create empirical
# counterfactuals for all treated. In bw, there are empirical 
# caunterfactuals for all treated except for the 1250 to 1500 range where 
# there is no overlap. I include the low birthwight treated in my analysis
# because I don't want to piroritize bw in my analysis.
# Note, that if we were estimating ATC bw would be more problematic, 
# because there are no counterfactuals for many
# healthy controls. 
# To summarize, there exist empirical copunterfactuals 
# for all treated and therefore we are not violating asusmptions. 

```


```{r}
### Answer 4(c)

balance.function <- function(data, cov, weights){
  
  balance <- data.frame(covariate = numeric(0), mn1 = numeric(0), mn0 = numeric(0),mn1.m= numeric(0), mn0.m=numeric(0), diff= numeric(0), diff.m=numeric(0),ratio=numeric(0), ratio.m=numeric(0))

  for(i in cov){
    
    mean_unmatched_control <- round(mean(subset(data, treat == 0)[,i]),3)
    mean_unmatched_treated <- round(mean(subset(data, treat == 1)[,i]),3)
    
    mean_matched_control <- round(weighted.mean(data[data$treat==0, i], weights[data$treat==0]),3)
    mean_matched_treated <- round(weighted.mean(data[data$treat==1, i], weights[data$treat==1]),3)
    
    unmatched_mean_difference <- round(ifelse(1 == range(data[,i])[2], mean_unmatched_treated - mean_unmatched_control, (mean_unmatched_treated - mean_unmatched_control)/sd(subset(data, treat == 1)[,i])), 3)
    matched_mean_difference <- round(ifelse(1 == range(data[,i])[2], mean_matched_treated - mean_matched_control, (mean_matched_treated - mean_matched_control)/sd(subset(data, treat == 1)[,i])), 3)
    
    unmatched_sd_ratio <- round(ifelse(1 == range(data[,i])[2],0, sd(subset(data, treat == 0)[,i]) / sd(subset(data, treat == 1)[,i])),3)
    
    matched_sd_ratio <- ifelse(1 == range(data[,i])[2] , 0, round(sqrt(wtd.var(data[data$treat==0, i], weights[data$treat==0])) / sqrt(wtd.var(data[data$treat==1, i], weights[data$treat==1])),3))
    
    output <- list(i, mean_unmatched_treated, mean_unmatched_control,  mean_matched_treated, mean_matched_control, unmatched_mean_difference, matched_mean_difference, unmatched_sd_ratio, matched_sd_ratio)
    balance[nrow(balance) + 1,] <- output
#    print(output)
  }
return(balance)
}

first_balance <- balance.function(df, covariates, weight)
first_balance

```

(d) How do you interpret the resulting balance?  In particular what are your concerns with regard to covariates that are not well balanced (3-4 sentences at most).

```{r}
### Answer 4(d)

# I think the above balance is sufficient. I am worried about
# the high ratio of standard deviation of dayskidh which might 
# reflect the matching of many treated to same control. Other 
# variables that are concerning are work.dur, income, momage, and dayskidh. 
# Income could be a confounder and predict the parent treating the child
# to high quality care independent of the study, but the difference
# in incomes between the groups doesn't seem to be large enough to 
# restructure the dataset. 

```

```{r}

### 4(e) -- test
df.test <- subset(data.frame(ppvtr.36=hw4$ppvtr.36, treat=hw4$treat, subset(hw4, select=c("bw", "b.marr"))), bw<3000)
covariates.test <- colnames(df)[3:length(colnames(df))]
test.fit <- glm(treat ~ bw + b.marr,family=binomial(link="logit"), data=df)
pscores.test <- test.fit$fitted.values
matches.t <- matching(z=df$treat, score=pscores.test, replace=TRUE)
#weight.test <- ifelse(df.test$treat == 0, matches.test$cnts, 1)
weight.test <- c(rep(1,sum(df.test$treat==1)),matches.t$cnts)
temp <-balance.function(df.test, c("bw", "b.marr"), weights=weight.test)
temp
```

\begin{verbatim}
             mn1       mn0     mn1.m     mn0.m    diff  diff.m  ratio  ratio.m
bw      2008.648  2629.482  2008.648  2001.838  -2.191   0.024  1.175   1.044
b.marr     0.431     0.595     0.431     0.486  -0.164  -0.055  0.000   0.000
\end{verbatim}

```{r}
### Answer 5

# without replacement
matches <- matching(z=df$treat, score=pscores, replace=FALSE)
weight2 <- ifelse(matches$matched!=0, 1, 0)
balance2 <- balance.function(df, covariates, weight2)
#balance2

# using matchit
rownames(df) <- NULL
match <- matchit(data = df, formula = treat ~ bw + bwg + hispanic + black + b.marr + lths + hs +
ltcoll + work.dur + prenatal + booze + cig + sex + first + preterm + momage + dayskidh + income, method = "optimal")
weight3 <- rep(0,nrow(df)) 
weight3[as.numeric(match$match.matrix)] <- 1
weight3[df$treat==1] <- 1
balance3 <- balance.function(df, covariates, weight3)
#balance3

# different covariates. I add the indicator for states where the RCT 
# was conducted and I treat square of bw as a covariate. 
df <- data.frame(ppvtr.36=hw4$ppvtr.36, treat=hw4$treat, subset(hw4, select=c(bw, bwg, hispanic, black, b.marr, lths, hs, ltcoll, work.dur, prenatal, booze, cig, sex, first, preterm, momage, dayskidh, income, st9, st12, st25, st36, st42, st48, st53)))
df$bw2 <- (df$bw)^2
covariates_sq <- colnames(df)[3:length(colnames(df))]
fit <- glm(treat ~ bw + I(bw^2) + bwg + hispanic + black + b.marr + lths + hs +
ltcoll + work.dur + prenatal + booze + cig + sex + first + preterm + momage + dayskidh + income+ st9 + st12 + st25 + st36 + st42 + st48 + st53 ,
family=binomial(link="logit"), data=df)
#pscores <- predict(fit, type='response') #same as line below
pscores <- fit$fitted.values
matches <- matching(z=df$treat, score=pscores, replace=TRUE)
weight4 <- c(rep(1,sum(df$treat==1)),matches$cnts)
balance4 <- balance.function(df, covariates_sq, weight4)
balance4
```


```{r}
### Answer 6
fit <- glm(treat ~ bw + bwg + hispanic + black + b.marr + lths + hs +
ltcoll + work.dur + prenatal + booze + cig + sex + first + preterm + momage + dayskidh + income,
family=binomial(link="logit"), data=df)
#pscores <- predict(fit, type='response') #same as line below
pscores <- fit$fitted.values

pscores_c <- pscores[df$treat==0]
pscores_c <- pscores_c/mean(pscores_c)

weight_IPTW <- c(rep(1,sum(df$treat==1)), pscores_c)
balance_IPTW <- balance.function(df, covariates, weight_IPTW)
```


```{r}
### Answer 7
comp_table <- data.frame(first_balance[,c(7,9)], balance2[,c(7,9)], balance3[,c(7,9)], balance4[c(1:18),c(7,9)], balance_IPTW[,c(7,9)], row.names=first_balance$covariate)
colnames(comp_table) <- c("diff.m 1", "ratio.m 1", "diff.m 2", "ratio.m 2", "diff.m 3", "ratio.m 3", "diff.m 4" , "ratio.m 4", "diff.m IPTW", "ratio.m IPTW")
comp_table

# I would choose my first propensity score model
# or IPTW because they have the best balance. First
# model has better balance but matches many control units 
# to one treated unit. As I explained in 4(c) the difference 
# in means and ratio of standard deviations remains big for income
# dayskidh and momage but they are still better than the other models. 
# 

```


```{r}
### Answer 8
effect1 <- lm(ppvtr.36 ~ treat, data=df, weights = weight1)$coefficients[2]
effect2 <- lm(ppvtr.36 ~ treat, data=df, weights = weight2)$coefficients[2]
effect3 <- lm(ppvtr.36 ~ treat, data=df, weights = weight3)$coefficients[2]
effect4 <- lm(ppvtr.36 ~ treat, data=df, weights = weight4)$coefficients[2]
#effect5 <- lm(ppvtr.36 ~ treat, data=df, weights = weight5)$coefficients[2]
effect_IPTW <- lm(ppvtr.36 ~ treat, data=df, weights = weight_IPTW)$coefficients[2]
treatment_effect <- c(effect1, effect2, effect3, effect4, effect_IPTW)
names(treatment_effect) <- c("effect1", "effect2","effect3","effect4", "effect_IPTW")
treatment_effect
```

```{r}
### Answer 9

# (1) ignorability
## We have measured all confounders and therefore we can 
## ignore the effects of the unobserved factors. 

# (2) sufficient overlap (positivity)
## We can make inferences about treatment effect on the 
## treated or controls only  over the area of common 
## support. We have satisfied this asusmption when we 
## have empirical counterfactuals for all treated (for ATT) 
## or all controls (for ATC)

# (3) appropriate specification of the propensity score model
## We know the matching based on the propensity score 
## model is appropriate when there is balance between 
## the comparison groups

# (4) Stable Unit Treatment Value Assumption
## The assumption that the effect of treatment is independent 
## of the composition of treatment selection

# Parametric (5)
## On the area of common support linearity holds. 

```


```{r}
### Answer 10

# Causal interpretation for my original propensity score estimate (effect1):
# The average treatment effect on the treated (ATT) was 6.26 IQ 
# points. In the counterfactual case without treatment, we would 
# observe IQ scores 6.26 points lower on average after three years. 
# We conclude that the treatment had causally contributed to a 
# significant increase in children's IQ scores. 

```

### Question 11: Comparison to linear regression 
Fit a regression of your outomes to the treatment indicator and covariates.
(a)  Report your estimate and standard error.
(b)  Interpret your results non-causally.
(c)  Why might we prefer the results from the propensity score approach to the linear regression results in terms of identifying a causal effect?

```{r}
### Answer 11
fit <- lm(ppvtr.36 ~ treat + bw + bwg + hispanic + black + b.marr + lths + hs +
ltcoll + work.dur + prenatal + booze + cig + sex + first + preterm + momage + dayskidh + income, data=df)
# (a)
summary(fit)$coefficients[2,][1:2]

# (b) 
## Treatment is correlated with an IQ test score increase of 10.98.
# Children who received treatment score better on the IQ test by an average of 10.98.

# (c)
## Linear regression violates the overlap assumption
## and therefore cannot be used to identify causal effects. 
## The treatment and control groups are not sufficiently simillar
## to one another to warrant counterfactual inferences. 

```


